---
title: "Individual Hypothesis Test"
format: html
editor: visual
---

## Individual Hypothesis Test

Null Hypothesis:

-   There is no significant difference in mortality between individuals within the age range of 25 to 29 years old and those who are greater than sixty-five years old from HIV in the years 2002 and 2013.​

Alternative Hypothesis:

-   Individuals in the age group range of 25 to 29 years old have suffered lower rates of mortality than those who are greater than sixty-five years old from HIV for the years 2002 and 2013.​​

```{r}

#Load the packages for the data analysis and visualization

library(dplyr)
library(ggplot2)
library(sqldf)
library(tidyr)
library(infer) # for hypothesis testing
library(countrycode)

```

```{r}
# Read the data that was located on Desktop 

HIV_TB <- read.csv("Desktop/HIV_TB_INCIDENCE_PREVALENCE_DEATHS.csv")
Malaria <- read.csv("Desktop/MALARIA_INCIDENCE_DEATHS.csv")

valid_iso3 <- unique(countrycode::codelist$iso3c) # checks for valid iso3 codes

# Set up the necessary variables to perform the hypothesis testing for the difference in the mean mortality rate for the years 2002 and 2013. 

HIV_TB_young_2002 <- HIV_TB %>%
  filter(age_group_name == "25 to 29",
         location_code %in% valid_iso3,
         metric == "Deaths",
         unit == "Number",
         sex_name == "Both sexes",
         cause_name == "HIV/AIDS",
         year == 2002) %>%
         pull(mean)

HIV_TB_older_2002 <- HIV_TB %>%
  filter(age_group_name %in% c("65 to 69", "70 to 74", "80 plus"),
         location_code %in% valid_iso3,
         metric == "Deaths",
         unit == "Number",
         sex_name == "Both sexes",
         year == 2002) %>%
         pull(mean)

HIV_TB_young_2013 <- HIV_TB %>%
  filter(age_group_name == "25 to 29",
         location_code %in% valid_iso3,
         metric == "Deaths",
         unit == "Number",
         sex_name == "Both sexes",
         cause_name == "HIV/AIDS",
         year == 2013) %>%
         pull(mean)

HIV_TB_older_2013 <- HIV_TB %>%
  filter(age_group_name %in% c("65 to 69", "70 to 74", "80 plus"),
         location_code %in% valid_iso3,
         metric == "Deaths",
         unit == "Number",
         sex_name == "Both sexes",
         cause_name == "HIV/AIDS", 
         year == 2013) %>%
         pull(mean)

# Converting to numeric to handle any potential non-numeric values
HIV_TB_young_numeric_2002 <- as.numeric(as.character(HIV_TB_young$mean))
HIV_TB_older_numeric_2002 <- as.numeric(as.character(HIV_TB_older$mean))
HIV_TB_young_numeric_2013 <- as.numeric(as.character(HIV_TB_young$mean))
HIV_TB_older_numeric_2013 <- as.numeric(as.character(HIV_TB_older$mean))

# Remove any NA values
HIV_TB_young_numeric_2002 <- HIV_TB_young_numeric[!is.na(HIV_TB_young_numeric)]
HIV_TB_older_numeric_2002 <- HIV_TB_older_numeric[!is.na(HIV_TB_older_numeric)]
HIV_TB_young_numeric_2013 <- HIV_TB_young_numeric[!is.na(HIV_TB_young_numeric)]
HIV_TB_older_numeric_2013 <- HIV_TB_older_numeric[!is.na(HIV_TB_older_numeric)]

# Perform t-test for the year 2002 for younger and older population.
HIV_2002_test <- t.test(
  x = HIV_TB_young_numeric_2002, 
  y = HIV_TB_older_numeric_2002, 
  alternative = "two.sided",
  paired = FALSE, 
  conf.level = 0.95
)

# Perform t-test for the year 2013 for younger and older population.
HIV_2013_test <- t.test(
  x = HIV_TB_young_numeric_2013, 
  y = HIV_TB_older_numeric_2013, 
  alternative = "two.sided",
  paired = FALSE,
  conf.level = 0.95
)

HIV_2002_test
HIV_2013_test

```

## Visualization Process 

```{r}

plot_data <- data.frame(
  year = rep(c(2002, 2013), each = 2),
  age_group = rep(c("20-25", "65 and older"), times = 2),
  mean_mortality = c(
    mean(HIV_TB_young_numeric_2002),
    mean(HIV_TB_older_numeric_2002),
    mean(HIV_TB_young_numeric_2013),
    mean(HIV_TB_older_numeric_2013)
  )
)

ggplot(plot_data, aes(x = factor(year), y = mean_mortality, fill = age_group)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  labs(
    title = "Mortality Rates Comparison",
    x = "Age Group",
    y = "Mean Mortality Rate",
    fill = "Age Group"
  ) +
  theme_minimal()
```
